Online supplement accompanying 'The Locally Gaussian Partial Correlation

### 2020-09-16 lgpc_JBES_supp.Rmd
New set of comments from Dag

We proceed now to establish the notion of conditional independence between $\Z_1$ and $\Z_2$ using the LGPC in the multivariate case. The conditional densities of $\Z_1|\Z_3$ and $\Z_2|\Z_3$ in terms of the local Gaussian parameters are given by

@@

\begin{equation}
\fSigma_{1|3} = \fSigma_{11}-\fSigma_{13}\fSigma_{33}^{-1}\fSigma_{31}, \,\, \fSigma_{2|3} = \fSigma_{22}-\fSigma_{23}\fSigma_{33}^{-1}\fSigma_{32}.
(\#eq:localsigma)
\end{equation}
Under conditional independence between $\Z_1$ and $\Z_2$ given $\Z_3$, the joint conditional density of $(\Z_1, \Z_2|\Z_3)$ is given by

@@

\begin{itemize}
\item[\textbf{a)}] \textbf{Let} $\bm{d_1 = d_2 = d_3 = 1}$. Consider the partial covariance matrix, where we again drop the $z$-dependence in all parameters: 
$$\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21}.$$
Here, since all variables are normalized with variance one, we have that
$$\fSigma = \begin{pmatrix} 1 & \rho_{12} & \rho_{13} \\ \rho_{12} & 1 & \rho_{23} \\ \rho_{31} & \rho_{32} & 1 \end{pmatrix},$$
and 
$$\fSigma^{11} = \begin{pmatrix} 1 & \rho_{12} \\ \rho_{21} & 1 \end{pmatrix}, \,\, \fSigma^{12} = \begin{pmatrix} \rho_{13} \\ \rho_{23} \end{pmatrix},\,\, \left(\fSigma^{22}\right)^{-1} = 1, \,\, \fSigma^{21} = \begin{pmatrix} \rho_{31} & \rho_{32} \end{pmatrix}.$$
Hence,
\begin{align*}
\fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21} &=  \begin{pmatrix} 1 & \rho_{12} \\ \rho_{21} & 1 \end{pmatrix} - \begin{pmatrix} \rho_{13} \\ \rho_{23} \end{pmatrix} \begin{pmatrix} \rho_{31} & \rho_{32} \end{pmatrix} \\
&= \begin{pmatrix} 1 - \rho_{13}^2 & \rho_{12} - \rho_{13}\rho_{23} \\ \rho_{12} - \rho_{13}\rho_{23} & 1 - \rho_{23}^2 \end{pmatrix}.
\end{align*}
This is a covariance matrix, and not a correlation matrix, so normalizing with standard deviations results in the partial correlation matrix
\begin{equation}
\begin{pmatrix}
1 & \frac{\rho_{12} - \rho_{13}\rho_{23}}{\sqrt{1-\rho_{13}^2}\sqrt{1-\rho_{23}^2}} \\  \frac{\rho_{12} - \rho_{13}\rho_{23}}{\sqrt{1-\rho_{13}^2}\sqrt{1-\rho_{23}^2}} & 1 
\end{pmatrix}.
\label{eq:partialmatrix}
\end{equation}
This result can also be obtained by direct application of \eqref{eq:matrixdef}:
$$\begin{pmatrix}1-\rho_{13}^2 & 0 \\ 0 & 1-\rho_ {23}^2\end{pmatrix}^{-1/2} \begin{pmatrix}  1 - \rho_{13}^2 & \rho_{12} - \rho_{13}\rho_{23} \\ \rho_{12} - \rho_{13}\rho_{23} & 1 - \rho_{23}^2  \end{pmatrix} \begin{pmatrix}1-\rho_{13}^2 & 0 \\ 0 & 1-\rho_ {23}^2\end{pmatrix}^{-1/2}.$$

@@

For fixed bandwidths we have from standard arguments, such as those provided by @hjort1996locally, that 
\begin{equation}
\sqrt{nb^3}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(0, \Jb^{-1}\Mb(\Jb^{-1})^T),
(\#eq:fixed-h)
\end{equation}
which corresponds to the usual rate for nonparametric density estimation. As $\hh\rightarrow0$, however, we must take extra care when considering the asymptotic covariance matrix. @tjostheim2013local, Section 4,  write the Taylor expansions of $\Mb$ and $\Jb$, now being functionals of three variables $(x_1, x_2, x_3)$ or $(z_1,z_2,z_3)$, as
$$\Mb = I_{\bm{M}} + II_{\bm{M}} + o(b^2) \,\,\, \textrm{ and } \Jb = I_{\bm{J}} + II_{\bm{J}} + o(b^2),$$
and consider each term. For example, we write their first term in the expansion of $\Mb$ in the three-variate case as follows:
$$I_{\bm{M}} \sim \int K^2(w_1, w_2, w_3)Ab_wb_w^TA^Tf(\z + b_1w_1 + b_2w_2 + b_3w_3)\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3,$$
where $K$ is a product kernel, and, in our case, $b_w$ is the vector defined by $b_w^T = \begin{pmatrix} 1 & b_1w_1 & b_2w_2 & b_3w_3 \end{pmatrix}$, and $A$ is the $3\times 4$ matrix $A = \begin{pmatrix} v & v_{z_1} & v_{z_2} & v_{z_3}\end{pmatrix}$ with $\fv^T =  \begin{pmatrix} v_1 & v_2 & v_3 \end{pmatrix}$, $v_i = \partial L_n/\partial \rho_i$, where the index $i=1,2,3$ represents the three local correlations, and $v_{z_i} = \partial \fv/\partial z_i$. In the next step, they compute the matrix

\noindent $\int K^2(w_1, w_2, w_3)b_wb_w^T\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3$, which, by omitting all constant factors and exploiting that $\int w^k K^2(w)\,\textrm{d}w = 0$ for $k = 1,3$, in our case becomes the diagonal matrix 
$$H = \begin{pmatrix} 1 & 0 & 0 & 0 \\ 0 & b_1^2w_1^2 & 0 & 0 \\ 0 & 0 & b_2^2w_2^2 & 0 \\ 0 & 0 & 0 & b_3^2w_3^2 \end{pmatrix}.$$

@@

\begin{align*}
\frac{1}{n}\sum_{t=1}^n \left(F_n(X_{i,t}) - F(X_{i,t})\right) &= 
\frac{1}{n^2}\sum_{t=1}^n\sum_{s \neq t}^n \left(I(X_{i,s} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{t=1}^n \left(1 - F(X_{i,t}\right) \\
&= \frac{1}{n} \sum_{t=1}^n \left(0.5 - F(X_{i,t})\right) + O_p\left(n^{-1}\right) \\
&= O_p\left(n^{-1/2}\right) + O_p\left(n^{-1}\right) \\
&= O_p\left(n^{-1/2}\right).
\end{align*}

@@

\begin{thm}
Let assumptions A1, A2 and A4 hold, and  let $\{L^{(b)}((Z_{i,t},Z_{j,t}),\rho)\}$, $\rho \in \Theta$ be Lipschitz continuous and 4-dominated on $\Theta$ uniformly in $t,n$. Moreover, assume that, as $b=b_n \to 0, \sigma_b^2 \stackrel{\textrm{def}}{=} \Var(L^{(b)}((Z_{i,t},Z_{j,t}),\rho) \to \sigma^2 < \infty$ and $\mu_b \stackrel{\textrm{def}}{=} \E(L^{(b)}(Z_{i,t},Z_{j,t}),\rho)) \to \mu < \infty$, for all $t=1,\ldots,n$ and $\rho \in \Theta$. 
Then,
\begin{equation}
\rho_{n,b}^* - \rho_{b,n} \stackrel {\mathbb{P}_{\omega}^{*},\mathbb{P}}{\to} 0.
\label{eq:thmB}
\end{equation}
\label{thm:E1}
\end{thm}

@@

Theorems \@ref(thm:E1) and \@ref(thm:E2) are formulated in their simplest case here, which concerns only one local parameter. A more complicated version is given by @lacal2018estimating concerning the five parameter case. The proofs are essentially the same and are spelled out in some detail in the supplementary material of that paper as Theorems D1 and D2.

@@

Let $\M_{ij}$ and $\J_{ij}$ be the $M$-matrix and $J$-matrix appearing in the proof of Corollary \ref{cor:E4} for a fixed $\rho_{ij}$. Note that in the one-parameter case of $\rho_{ij}$ in Corollary \ref{cor:E4}, to first order the covariance matrix $\fLambda$ is a diagonal matrix having ordered values of $M_{ij}/J_{ij}^2$ along the diagonal. This can be seen using the methods of proofs of Theorem 2 of @otneim2017locally and Theorem 2 @otneim2017conditional. Note also that the 3-coordinate case of alternative (i) can be stated and proved in the same way, but a different scaling factor is required to avoid a singular covariance matrix as seen in the proof of Theorem 3.1 in the main article. The 3-coordinate case is in fact closer to the situation in @lacal2018estimating.

@@

Consider the general case $\Z = (\Z_1, \Z_2, \Z_3)$ where each component is $d_i$-dimensional correspondingly, with $i = 1,2,3$, and $\Z$, naturally, is $d_1+d_2+d_3$-dimensional, and consider also the local partial correlation matrix of $(\Z_1,\Z_2|\Z_3)$. In alternative (iii), where $d_3 = 1$, in Section \@ref(chap:estimation) this matrix is given by equation \@ref(eq:offdiag3) with
$$
\alpha_{ij} = \rho(Z_{1,i},Z_{2,j}|Z_3), \;\; i=1,\ldots,d_1; j = 1,\dots,d_2.
$$
This forms a $d_1 \times d_2$ matrix ${\falpha}$. One can test for conditional independence between $Z_{1,i}$ and $Z_{2,j}$ given $\Z_3$ by testing whether $\alpha_{ij}=0$, or one can test for conditional independence between $\Z_1$ and $\Z_2$ given $\Z_3$ by testing whether ${\falpha} = \{\alpha_{ij}\} = {\bm{0}}$. The latter case is in a sense the most general, and we will stick to that in the sequel. One can then test for ${\falpha} = {\bm{0}}$ by using a matrix norm, for instance the Frobenius norm, where $||{\falpha}|| = \sqrt{\sum_{i,j} \alpha_{ij}^2} = 0$. A more convenient test statistic is $\sum_{i,j} \alpha_{ij}^2$. To follow the notation of @lacal2018estimating we write, letting $\z=(\z_1,\z_2,\z_3)$,
$$
T = \int_S h(\frho(\z)) \di F(\z)
$$
where $S$ is an a area of integration, $F$ is the cumulative distribution function of $\Z$, and $\frho = \frho(\z)$, as before, is a vector consisting of all possible pairs that can be formed by picking pairwise components from $\Z$. The number of components of this vector depends on which of the alternatives models (i)--(v) in Section \@ref(chap:estimation) one considers. It is reasonable to express $T$ as a function of $\frho$ since all partial correlations can be expressed as a function of this vector (see e.g. again equation \@ref(eq:offdiag3)). One possible choice of the function $h$ is $h(\frho(\z)) = \sum_{i,j} \alpha_{i,j}^2(z_{1,i},z_{2,j}|\ztwo)$, which is a function of $\frho$ since the $\alpha$-s are functions of $\frho$. The corresponding function in @lacal2018estimating is $h(\theta(x))$, where $\theta(x)$ is at most 5-dimensional and $\x$ is bivariate. As in that paper we proceed in two stages. First we let the bandwidth vector $\bb$ be fixed and let $n \to \infty$, and then we let $\bb = (b_1,b_2) = (b,b) \to \bm{0}$.
Accordingly we define
$$
T_b = \int_S h(\frho_b(\z)) \di F(\z) 
$$
where $\frho_b$ is the population quantity for a fixed bandwidth $\bb$, and $T_{n,b}$ by
$$
T_{n,b} = \int_S h(\frho_{b,n}(\z)) \di F_n(\z),
$$
where $\frho_{b,n}(\z)$ is the estimate of $\frho(\z)$ obtained by pairwise likelihood for each component of $\frho(\z)$.

@@

It is more difficult to obtain the limiting distribution. We will follow the technique in the proof of Theorem 3.1 in @lacal2018estimating. See also @joe1989estimation. The details are a bit more involved in the present case because of the larger dimension and because we need to involve the pairwise log likelihood in the procedure.

@@

\begin{thm} 
Let the assumptions A, D--F of Theorem 3 of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) and the conditions of Theorem 3.2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) (which includes conditions A1 and A2 of this supplement) be fulfilled. Then,
as $n \to \infty$,
$$
\sqrt{n}\big(C_{n}(A_b)\big)^{-\frac{1}{2}}(T_{n,b} - T_b) \buildrel {d} \over {\to} {\cal N}(0,1),
$$
where
\begin{align*}
A_b(\z)  & = \int_{S} \left[\sum_{i,j}\left\{\frac{\partial}{\partial \rho_{ij}} h(\frho_b,(\fv)\right\}J_{ij}^{-1}(\rho_{ij,b}(\fv_{ij}))K_b(\z_{ij}- \fv_{ij})u_{ij}(\z_{ij},\rho_{ij,b}(\fv_{ij}))\right] \di F(\fv) \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad + h(\frho_b(\z))1_S(\z)
\end{align*}
and 
\begin{align}
C_n(A_b) &= \int A_b^2(\z) \di F(\z) + \frac{1}{n}\sum_{s \neq t}\int A_b(\z)A_b(\fu) \di F^{s,t}(\z,\fu) \nonumber \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad - n\int A_b(\z)A_b(\fu) \di F(\z) \di F(\fu)
(\#eq:t0)
\end{align}
with $J_{ij}$ being the bivariate version of $\J_b$ defined in Theorem 3.1, and, with $F^{s,t}(\z,\fu)$ being the joint cumulative distribution function of $(\Z_s,\Z_t)$, and where $\{\Z_{t}, t=1,\ldots,n\}$ are the $n$ observations of $\Z$ and $u_{ij}(\Z_{ij}, \rho_{ij,b}(\fv_{ij})) = \frac{\partial}{\partial \rho_{ij,b}}\log\psi(\Z_{ij},\rho_{ij,b}(\fv_{ij}))$.

@@

Let $(i,j)$ be a fixed pair and $\rho_{ij,n,b}(\z_{ij})$ be the corresponding estimated local correlation for the pair $(i,j)$ chosen from the vector $\frho_{n,b}$. Here the pair of coordinates $\z_{ij} = (z_i,z_j)$ corresponds to the bivariate $\x$ in the proof of Theorem 3.1 by @lacal2017local. We will also omit some indices or parts of indices when it is clear what is meant. Finally, we omit the area of integration $S$. We are using the pairwise local log likelihood given by
$$
L_{ij,n} = \frac{1}{n} \sum_{t=1}^{n} K_b(\Z_{ij,t}-\z_{ij}) \log \psi\left(\Z_{ij},\rho_{b,ij}(\z_{ij})\right) - \int K_b(\fv_{ij}-\z_{ij}) \psi_{ij}(\fv_{ij},\rho_{b,ij}(\z_{ij})) \di \fv_{ij},
$$
where $K_b = b_1^{-1}K_1(\cdot/b_1) b_2^{-1} K_2(\cdot/b_2)$ is a product kernel function, and $\Z_{ij,t} = (Z_{i,t},Z_{j,t})$ is the $t$-th pair of observations, and $\fv_{ij} = (v_i,v_j)$. Estimates $\rho_{ij,b,n}(\z_{ij})$ are obtained by differentiation of the log likelihood and by introducing the score function $U_{ij}$ and information function $I_{ij}$ as follows
\begin{align*}
U_{ij}(\z_{ij},\rho_{ij}) & = \frac{\partial}{\partial \rho_{ij}}L_{ij,n}(\z_{ij},\rho_{ij}) \\
& = \frac{1}{n}\sum_{t=1}^{n}K_b(\Z_{ij,t}-\z_{ij})u_{ij}(\Z_{ij,t},\rho_{ij}) - 
\int K_b(\fv_{ij}-\z_{ij})u(\fv_{ij},\rho_{ij})\psi(\fv_{ij},\rho_{ij}) \di \fv_{ij}
\end{align*}
where $u_{ij}(\z_{ij},\rho_{ij}) = \partial/\partial \rho_{ij} \log \psi(\z_{ij},\rho_{ij})$. Similarly
$$
I_{ij}(\z_{ij},\rho_{ij}) = \frac{\partial^2}{\partial \rho_{ij}^2} L_{ij,n}(\z_{ij}, \rho_{ij}) \quad \mbox{and} \quad J_{ij,b}(\z_{ij}) = \E(-I_{ij}(\z_{ij},\rho_{ij})),
$$
where Lacal and Tjøstheim [-@lacal2017local; -@lacal2018estimating] may be consulted for more details in the derivation and a more explicit expression for $I_{ij}(\z_{ij},\rho_{ij})$ for the case considered there. In contradistinction to the 5-parameter case considered by @lacal2018estimating, in our present 1-parameter case $J_{b,ij}$ is not singular as $\bb \to \bm{0}$. In fact, it is shown by Otneim and Tjøstheim [-@otneim2017locally, Theorem 2] that as $\bb \to \bm{0}$, $J_{b,ij} \to u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho(\z_{ij}))$.

>>>>

We proceed now to establish the notion of conditional independence between $\Z_1$ and $\Z_2$ using the LGPC in the multivariate case. The conditional densities of $\Z_1|\Z_3$ and $\Z_2|\Z_3$ in terms of the local Gaussian parameters at the point $(\z_1, \z_2, \z_3)$ are given by

@@

\begin{equation}
\fSigma_{1|3} = \fSigma_{11}-\fSigma_{13}\fSigma_{33}^{-1}\fSigma_{31}, \,\, \fSigma_{2|3} = \fSigma_{22}-\fSigma_{23}\fSigma_{33}^{-1}\fSigma_{32}.
(\#eq:localsigma)
\end{equation}
Under conditional independence between $\Z_1$ and $\Z_2$ given $\Z_3$, the joint conditional density of $(\Z_1, \Z_2|\Z_3)$ at the point $(\z_1, \z_2, \z_3)$ is given by

@@

\begin{itemize}
\item[\textbf{a)}] \textbf{Let} $\bm{d_1 = d_2 = d_3 = 1}$. Consider the partial covariance matrix, where we again drop the $z$-dependence in all parameters: 
$$\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21}.$$
Here, since all variables are normalized with variance one, we have that
$$\fSigma = \begin{pmatrix} 1 & \rho_{12} & \rho_{13} \\ \rho_{21} & 1 & \rho_{23} \\ \rho_{31} & \rho_{32} & 1 \end{pmatrix},$$
and 
$$\fSigma^{11} = \begin{pmatrix} 1 & \rho_{12} \\ \rho_{21} & 1 \end{pmatrix}, \,\, \fSigma^{12} = \begin{pmatrix} \rho_{13} \\ \rho_{23} \end{pmatrix},\,\, \left(\fSigma^{22}\right)^{-1} = 1, \,\, \fSigma^{21} = \begin{pmatrix} \rho_{31} & \rho_{32} \end{pmatrix}.$$
Hence,
\begin{align*}
\fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21} &=  \begin{pmatrix} 1 & \rho_{12} \\ \rho_{21} & 1 \end{pmatrix} - \begin{pmatrix} \rho_{13} \\ \rho_{23} \end{pmatrix} \begin{pmatrix} \rho_{31} & \rho_{32} \end{pmatrix} \\
&= \begin{pmatrix} 1 - \rho_{13}^2 & \rho_{12} - \rho_{13}\rho_{23} \\ \rho_{12} - \rho_{13}\rho_{23} & 1 - \rho_{23}^2 \end{pmatrix}.
\end{align*}
This is a covariance matrix, and not a correlation matrix, so normalizing with standard deviations results in the partial correlation matrix
\begin{equation}
\begin{pmatrix}
1 & \frac{\rho_{12} - \rho_{13}\rho_{23}}{\sqrt{1-\rho_{13}^2}\sqrt{1-\rho_{23}^2}} \\  \frac{\rho_{12} - \rho_{13}\rho_{23}}{\sqrt{1-\rho_{13}^2}\sqrt{1-\rho_{23}^2}} & 1 
\end{pmatrix}.
\label{eq:partialmatrix}
\end{equation}
This result can also be obtained by direct application of \eqref{eq:matrixdef}:
$$\begin{pmatrix}1-\rho_{13}^2 & 0 \\ 0 & 1-\rho_ {23}^2\end{pmatrix}^{-1/2} \begin{pmatrix}  1 - \rho_{13}^2 & \rho_{12} - \rho_{13}\rho_{23} \\ \rho_{12} - \rho_{13}\rho_{23} & 1 - \rho_{23}^2  \end{pmatrix} \begin{pmatrix}1-\rho_{13}^2 & 0 \\ 0 & 1-\rho_ {23}^2\end{pmatrix}^{-1/2}.$$

@@

For fixed bandwidths we have from standard arguments, such as those provided by @hjort1996locally, that 
\begin{equation}
\sqrt{nb^3}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(0, \Jb^{-1}\Mb(\Jb^{-1})^T),
(\#eq:fixed-h)
\end{equation}
which corresponds to the usual rate for nonparametric density estimation. As $\hh\rightarrow0$, however, we must take extra care when considering the asymptotic covariance matrix. @tjostheim2013local, Section 4,  write the Taylor expansions of $\Mb$ and $\Jb$, now being functionals of three variables $(x_1, x_2, x_3)$ or $(z_1,z_2,z_3)$, as
$$\Mb = I_{\bm{M}} + II_{\bm{M}} + o(b^2) \,\,\, \textrm{ and } \Jb = I_{\bm{J}} + II_{\bm{J}} + o(b^2),$$
and consider each term. For example, we write their first term in the expansion of $\Mb$ in the three-variate case as follows:
$$I_{\bm{M}} \sim \int K^2(w_1, w_2, w_3)Ab_wb_w^TA^Tf(\z + b_1w_1 + b_2w_2 + b_3w_3)\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3,$$
where $K$ is a product kernel, and, in our case, $b_w$ is the vector defined by $b_w^T = \begin{pmatrix} 1 & b_1w_1 & b_2w_2 & b_3w_3 \end{pmatrix}$, and $A$ is the $3\times 4$ matrix $A = \begin{pmatrix} v & v_{z_1} & v_{z_2} & v_{z_3}\end{pmatrix}$ with $\fv^T =  \begin{pmatrix} v_1 & v_2 & v_3 \end{pmatrix}$, $v_i = \partial L_n/\partial \rho_i$, where the index $i=1,2,3$ represents the three local correlations, and $v_{z_i} = \partial \fv/\partial z_i$. In the next step, they Taylor expand $f$ and compute the matrix $\int K^2(w_1, w_2, w_3)b_wb_w^T\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3$, which, by omitting all constant factors and exploiting that $\int w^k K^2(w)\,\textrm{d}w = 0$ for $k = 1,3$, in our case becomes the diagonal matrix 
$$H = \begin{pmatrix} 1 & 0 & 0 & 0 \\ 0 & b_1^2w_1^2 & 0 & 0 \\ 0 & 0 & b_2^2w_2^2 & 0 \\ 0 & 0 & 0 & b_3^2w_3^2 \end{pmatrix}.$$

@@

\begin{align*}
\frac{1}{n}\sum_{t=1}^n \left(F_n(X_{i,t}) - F(X_{i,t})\right) &= 
\frac{1}{n^2}\sum_{t=1}^n\sum_{s \neq t}^n \left(I(X_{i,s} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{t=1}^n \left(1 - F(X_{i,t})\right) \\
&= \frac{1}{n} \sum_{t=1}^n \left(0.5 - F(X_{i,t})\right) + O_p\left(n^{-1}\right) \\
&= O_p\left(n^{-1/2}\right) + O_p\left(n^{-1}\right) \\
&= O_p\left(n^{-1/2}\right).
\end{align*}

@@

\begin{thm}
Let assumptions A1, A2 and A4 hold, and  let $\{L^{(b)}((Z_{i,t},Z_{j,t}),\rho)\}$, $\rho \in \Theta$ be Lipschitz continuous and 4-dominated on $\Theta$ uniformly in $t,n$. Moreover, assume that, as $b=b_n \to 0, \sigma_b^2 \stackrel{\textrm{def}}{=} \Var(L^{(b)}((Z_{i,t},Z_{j,t}),\rho)) \to \sigma^2 < \infty$ and $\mu_b \stackrel{\textrm{def}}{=} \E(L^{(b)}(Z_{i,t},Z_{j,t}),\rho)) \to \mu < \infty$, for all $t=1,\ldots,n$ and $\rho \in \Theta$. 
Then,
\begin{equation}
\rho_{n,b}^* - \rho_{b,n} \stackrel {\mathbb{P}_{\omega}^{*},\mathbb{P}}{\to} 0.
\label{eq:thmB}
\end{equation}
\label{thm:E1}
\end{thm}

@@

Theorems \@ref(thm:E1) and \@ref(thm:E2) are formulated in their simplest case here, which concerns only one local parameter. A more complicated version is given by @lacal2018estimating concerning the five parameter case. The proofs are essentially the same and are spelled out in some detail in the supplementary material of that paper as the proofs of Theorems D1 and D2.

@@

Let $\M_{ij}$ and $\J_{ij}$ be as in the proof of Corollary \ref{cor:E4} for a fixed $\rho_{ij}$. Note that in the one-parameter case of $\rho_{ij}$ in Corollary \ref{cor:E4}, to first order the covariance matrix $\fLambda$ is a diagonal matrix having ordered values of $M_{ij}/J_{ij}^2$ along the diagonal. This can be seen using the methods of proofs of Theorem 2 of @otneim2017locally and Theorem 2 @otneim2017conditional. Note also that the 3-coordinate case of alternative (i) can be stated and proved in the same way, but a different scaling factor is required to avoid a singular covariance matrix as seen in the proof of Theorem 3.1 in the main article. The 3-coordinate case is in fact closer to the situation in @lacal2018estimating.

@@

Consider the general case $\Z = (\Z_1, \Z_2, \Z_3)$ where each component is $d_i$-dimensional correspondingly, with $i = 1,2,3$, and $\Z$, naturally, is $d_1+d_2+d_3$-dimensional, and consider also the local partial correlation matrix of $(\Z_1,\Z_2|\Z_3)$. In alternative (iii), where $d_3 = 1$, in Section \@ref(chap:estimation) this matrix is given by equation \@ref(eq:offdiag3) with
$$
\alpha_{ij} = \rho(Z_{1,i},Z_{2,j}|Z_3), \;\; i=1,\ldots,d_1; j = 1,\dots,d_2.
$$
This forms a $d_1 \times d_2$ matrix ${\falpha}$. One can test for conditional independence between $Z_{1,i}$ and $Z_{2,j}$ given $\Z_3$ by testing whether $\alpha_{ij}=0$, or one can test for conditional independence between $\Z_1$ and $\Z_2$ given $\Z_3$ by testing whether ${\falpha} = \{\alpha_{ij}\} = {\bm{0}}$. The latter case is in a sense the most general, and we will stick to that in the sequel. One can then test for ${\falpha} = {\bm{0}}$ by using a matrix norm, for instance the Frobenius norm, where $||{\falpha}|| = \sqrt{\sum_{i,j} \alpha_{ij}^2} = 0$. A more convenient test statistic is $\sum_{i,j} \alpha_{ij}^2$. To follow the notation of @lacal2018estimating we write, letting $\z=(\z_1,\z_2,\z_3)$,
$$
T = \int_S h(\frho(\z)) \di F(\z)
$$
where $S$ is an a area of integration, $F$ is the cumulative distribution function of $\Z$, and $\frho = \frho(\z)$, as before, is a vector consisting of all possible pairs that can be formed by picking pairwise components from $\Z$. The number of components of this vector depends on which of the alternatives models (i)--(v) in Section \@ref(chap:estimation) one considers. It is reasonable to express $T$ as a function of $\frho$ since all partial correlations can be expressed as a function of this vector (see e.g. again equation \@ref(eq:offdiag3)). One possible choice of the function $h$ is $h(\frho(\z)) = \sum_{i,j} \alpha_{i,j}^2(z_{1,i},z_{2,j}|\ztwo)$, which is a function of $\frho$ since the $\alpha$-s are functions of $\frho$. The corresponding function in @lacal2018estimating is $h(\theta(\x), \x)$, where $\theta(\x)$ is at most 5-dimensional and $\x$ is bivariate. As in that paper we proceed in two stages. First we let the bandwidth vector $\bb$ be fixed and let $n \to \infty$, and then we let $\bb = (b_1,b_2) = (b,b) \to \bm{0}$.
Accordingly we define
$$
T_b = \int_S h(\frho_b(\z)) \di F(\z) 
$$
where $\frho_b$ is the population quantity for a fixed bandwidth $\bb$, and $T_{n,b}$ by
$$
T_{n,b} = \int_S h(\frho_{b,n}(\z)) \di F_n(\z),
$$
where $\frho_{b,n}(\z)$ is the estimate of $\frho(\z)$ obtained by pairwise likelihood for each component of $\frho(\z)$.

@@

It is more difficult to obtain the limiting distribution. We will follow the technique in the proof of Theorem 3.2 in @lacal2018estimating. See also @joe1989estimation. The details are a bit more involved in the present case because of the larger dimension and because we need to involve the pairwise log likelihood in the procedure.

@@

\begin{thm} 
Let the assumptions A, D--F of Theorem 3 of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) and the conditions of Theorem 3.2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) (which includes conditions A1 and A2 of this supplement) be fulfilled. Then,
as $n \to \infty$,
$$
\sqrt{n}\big(C_{n}(A_b)\big)^{-\frac{1}{2}}(T_{n,b} - T_b) \buildrel {d} \over {\to} {\cal N}(0,1),
$$
where
\begin{align*}
A_b(\z)  & = \int_{S} \left[\sum_{i,j}\left\{\frac{\partial}{\partial \rho_{ij}} h(\frho_b,(\fv))\right\}J_{ij}^{-1}(\rho_{ij,b}(\fv_{ij}))K_b(\z_{ij}- \fv_{ij})u_{ij}(\z_{ij},\rho_{ij,b}(\fv_{ij}))\right] \di F(\fv) \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad + h(\frho_b(\z))1_S(\z)
\end{align*}
and 
\begin{align}
C_n(A_b) &= \int A_b^2(\z) \di F(\z) + \frac{1}{n}\sum_{s \neq t}\int A_b(\z)A_b(\fu) \di F^{s,t}(\z,\fu) \nonumber \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad - n\int A_b(\z)A_b(\fu) \di F(\z) \di F(\fu)
(\#eq:t0)
\end{align}
with $J_{ij}$ being the bivariate version of $\J_b$ defined in Theorem 3.1, and, with $F^{s,t}(\z,\fu)$ being the joint cumulative distribution function of $(\Z_s,\Z_t)$, and where $\{\Z_{t}, t=1,\ldots,n\}$ are the $n$ observations of $\Z$ and $u_{ij}(\Z_{ij}, \rho_{ij,b}(\fv_{ij})) = \frac{\partial}{\partial \rho_{ij,b}}\log\psi(\Z_{ij},\rho_{ij,b}(\fv_{ij}))$.

@@

Let $(i,j)$ be a fixed pair and $\rho_{ij,n,b}(\z_{ij})$ be the corresponding estimated local correlation for the pair $(i,j)$ chosen from the vector $\frho_{n,b}$. Here the pair of coordinates $\z_{ij} = (z_i,z_j)$ corresponds to the bivariate $\x$ in the proof of Theorem 3.1 by @lacal2017local. We will also omit some indices or parts of indices when it is clear what is meant. Finally, we omit the area of integration $S$. We are using the pairwise local log likelihood given by
$$
L_{ij,n} = \frac{1}{n} \sum_{t=1}^{n} K_b(\Z_{ij,t}-\z_{ij}) \log \psi\left(\Z_{ij},\rho_{b,ij}(\z_{ij})\right) - \int K_b(\fv_{ij}-\z_{ij}) \psi_{ij}(\fv_{ij},\rho_{b,ij}(\z_{ij})) \di \fv_{ij},
$$
where $K_b = b_1^{-1}K_1(\cdot/b_1) b_2^{-1} K_2(\cdot/b_2)$ is a product kernel function, and $\Z_{ij,t} = (Z_{i,t},Z_{j,t})$ is the $t$-th pair of observations, and $\fv_{ij} = (v_i,v_j)$. Estimates $\rho_{ij,b,n}(\z_{ij})$ are obtained by differentiation of the log likelihood and by introducing the score function $U_{ij}$ and information function $I_{ij}$ as follows
\begin{align*}
U_{ij}(\z_{ij},\rho_{ij}) & = \frac{\partial}{\partial \rho_{ij}}L_{ij,n}(\z_{ij},\rho_{ij}) \\
& = \frac{1}{n}\sum_{t=1}^{n}K_b(\Z_{ij,t}-\z_{ij})u_{ij}(\Z_{ij,t},\rho_{ij}) - 
\int K_b(\fv_{ij}-\z_{ij})u_{ij}(\fv_{ij},\rho_{ij})\psi(\fv_{ij},\rho_{ij}) \di \fv_{ij}
\end{align*}
where $u_{ij}(\z_{ij},\rho_{ij}) = \partial/\partial \rho_{ij} \log \psi(\z_{ij},\rho_{ij})$. Similarly
$$
I_{ij}(\z_{ij},\rho_{ij}) = \frac{\partial^2}{\partial \rho_{ij}^2} L_{ij,n}(\z_{ij}, \rho_{ij}) \quad \mbox{and} \quad J_{ij,b}(\z_{ij}) = \E(-I_{ij}(\z_{ij},\rho_{ij})),
$$
where Lacal and Tjøstheim [-@lacal2017local; -@lacal2018estimating] may be consulted for more details in the derivation and a more explicit expression for $I_{ij}(\z_{ij},\rho_{ij})$ for the case considered there. In contradistinction to the 5-parameter case considered by @lacal2018estimating, in our present 1-parameter case $J_{b,ij}$ is not singular as $\bb \to \bm{0}$. In fact, it is shown by Otneim and Tjøstheim [-@otneim2017locally, Theorem 2] that as $\bb \to \bm{0}$, $J_{b,ij} \to u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho(\z_{ij}))$.
